---
title: "NLSY97 Analysis (Based on Gender)"
author: "Anum Malik, Rishabh Jain, Javier Rico Aristizabal"
date: ''
output: html_document
---

#### Package loading

```{r}
library(tidyverse)
library(knitr)
```


#### Importing the data

```{r}
# Import starting data
nlsy <- read_csv("http://www.andrew.cmu.edu/user/achoulde/94842/final_project/nlsy97/nlsy97_Nov2020.csv")
```

#### Variables present in the base data set

To learn more about the data, you can have a look at the [variable codebook file](http://www.andrew.cmu.edu/user/achoulde/94842/final_project/nlsy97/nlsy97_codebook.txt).


Here's how to rename all the variables to the Question Name abbreviation.  **You will want to change the names to be even more descriptive**, but this is a start.

```{r}
# Change column names to question name abbreviations (you will want to change these further)
colnames(nlsy) <- c("PSTRAN_GPA.01_PSTR",
    "INCARC_TOTNUM_XRND",
    "INCARC_AGE_FIRST_XRND",
    "INCARC_LENGTH_LONGEST_XRND",
    "PUBID_1997",
    "YSCH-36400_1997",
    "YSCH-37000_1997",
    "YSAQ-010_1997",
    "Marijuana_1997", #YSAQ-369_1997
    "YEXP-300_1997",
    "PerCh.Dip.20_1997", #YEXP-1500_1997
    "YEXP-1600_1997",
    "PerCh.Baby_1997", #YEXP-1800_1997
    "PerCh.Coll.Degree.30_1997", #YEXP-2000_1997
    "Sex",
    "BDATE_M_1997", #KEY_BDATE_M_1997
    "BDATE_Y_1997", #KEY_BDATE_Y_1997
    "PC8-090_1997",
    "PC8-092_1997",
    "PC9-002_1997",
    "Dep_F_1997", #PC12-024_1997
    "Dep_M_1997", #PC12-028_1997
    "CV_AGE_12/31/96_1997",
    "CV_BIO_MOM_AGE_CHILD1_1997",
    "CV_BIO_MOM_AGE_YOUTH_1997",
    "CV_CITIZENSHIP_1997",
    "CV_ENROLLSTAT_1997",
    "HH_NET_WORTH_P_1997", #CV_HH_NET_WORTH_P_1997
    "CV_YTH_REL_HH_CURRENT_1997",
    "CV_MSA_AGE_12_1997",
    "CV_URBAN-RURAL_AGE_12_1997",
    "CV_SAMPLE_TYPE_1997",
    "CV_HGC_BIO_DAD_1997",
    "CV_HGC_BIO_MOM_1997",
    "Highest.Grade_RES_DAD_1997", #CV_HGC_RES_DAD_1997
    "Highest.Grade_RES_MOM_1997", #CV_HGC_RES_MOM_1997
    "Race",
    "YSCH-6800_1998",
    "YSCH-7300_1998",
    "Ever_Used_Coca_1998", #YSAQ-372B_1998
    "YSAQ-371_2000",
    "Disorganized_2002", #YSAQ-282J_2002
    "YSAQ-282Q_2002",
    "HH_NET_WORTH_Y_2003", #CV_HH_NET_WORTH_Y_2003
    "CV_BA_CREDITS.01_2004",
    "YSAQ-000B_2004", 
    "YSAQ-373_2004",
    "Marijuana_2005", #YSAQ-369_2005
    "Children_HH_2007", #CV_BIO_CHILD_HH_2007
    "YTEL-52~000001_2007",
    "YTEL-52~000002_2007",
    "YTEL-52~000003_2007",
    "YTEL-52~000004_2007",
    "Children_HH_2009", #CV_BIO_CHILD_HH_2009
    "CV_COLLEGE_TYPE.01_2011",
    "Family_Income_2011", #CV_INCOME_FAMILY_2011
    "CV_HH_SIZE_2011",
    "CV_HH_UNDER_18_2011",
    "CV_HH_UNDER_6_2011",
    "Highest_Degree_Prior_2011.12", #CV_HIGHEST_DEGREE_1112_2011
    "Children_HH_2011", #CV_BIO_CHILD_HH_2011
    "Highest_Grade_Completed_2011", #YSCH-3112_2011
    "YSAQ-000A000001_2011", 
    "Height.Inch_2011", #YSAQ-000A000002_2011
    "Weight.Pound_2011", #YSAQ-000B_2011
    "YSAQ-360C_2011",
    "YSAQ-364D_2011",
    "YSAQ-371_2011", 
    "Hard.drugs.DLI_2011", #YSAQ-372CC_2011
    "YSAQ-373_2011",
    "YSAQ-374_2011",
    "Industry_Code_2011", #YEMP_INDCODE-2002.01_2011
    "Children_HH_2015", #CV_BIO_CHILD_HH_2015
    "Industry_Code_2017", #YEMP_INDCODE-2002.01_2017
    "Occupation_Code_2017", #YEMP_OCCODE-2002.01_2017
    "MaritalStatus_2017", #CV_MARSTAT_COLLAPSED_2017
    "YINC-1400_2017",
    "Income",
    "YINC-1800_2017", 
    "YINC-2400_2017",
    "YINC-2600_2017",
    "YINC-2700_2017",
    "CVC_YTH_REL_HH_AGE6_YCHR_XRND",
    "CVC_SAT_MATH_SCORE_2007_XRND",
    "CVC_SAT_VERBAL_SCORE_2007_XRND",
    "CVC_ACT_SCORE_2007_XRND",
    "HH_NET_WORTH_20", #CVC_HH_NET_WORTH_20_XRND
    "HH_NET_WORTH_25", #CVC_HH_NET_WORTH_25_XRND
    "ASSETS_FINANCIAL_25", #CVC_ASSETS_FINANCIAL_25_XRND
    "ASSETS_DEBTS_20", #CVC_ASSETS_DEBTS_20_XRND
    "HH_NET_WORTH_30", #CVC_HH_NET_WORTH_30_XRND
    "CVC_HOUSE_VALUE_30_XRND",
    "CVC_HOUSE_TYPE_30_XRND",
    "ASSETS_FINANCIAL", #CVC_ASSETS_FINANCIAL_30_XRND
    "ASSETS_DEBTS_30") #CVC_ASSETS_DEBTS_30_XRND

### Set all negative values to NA.  
### THIS IS DONE ONLY FOR ILLUSTRATIVE PURPOSES
### DO NOT TAKE THIS APPROACH WITHOUT CAREFUL JUSTIFICATION
nlsy[nlsy < 0]  <- NA
```

#### A note on missing values

Here's an example of what the variable description files look like

```
T76400.00    [YSAQ-372CC]                                   Survey Year: 2011
  PRIMARY VARIABLE

 
             HAS R USED COCAINE/HARD DRUGS SINCE DLI?
 
Excluding marijuana and alcohol, since the date of last interview, have you used
any drugs like cocaine, crack, heroin, or crystal meth, or any other substance 
not prescribed by a doctor, in order to get high or to achieve an altered state?
 
UNIVERSE: All except prisoners in an insecure environment
 
     215       1 YES   (Go To T76401.00)
    7023       0 NO
  -------
    7238
 
Refusal(-1)           74
Don't Know(-2)        26
TOTAL =========>    7338   VALID SKIP(-4)      85     NON-INTERVIEW(-5)    1561
 
Min:              0        Max:              1        Mean:                 .03
 
Lead In: T76397.00[Default] T76399.00[Default]  T76398.00[0:0]
Default Next Question: T76403.00
```

This description says that the numbers -1, -2, -4 and -5 all have a special meaning for this variable.  They denote different types of missingness.  You can recode all of these to `NA`, but you should also think about whether the different missigness indicators are in some way informative.  (i.e., if someone refuses to answer questions related to drug use, might this inform us about their income?) 

#### Getting to know our two main variables.

In the previous chunk of code we have appropriately renamed the variables corresponding to `sex`, `race` and `income` (as reported on the 2017 survey).  Let's have a quick look at what we're working with.

```{r}
table(nlsy$Sex)

table(nlsy$Race)
```

The data codebook tells us that the coding for sex is `Male = 1`, `Female = 2`.  For the race/ethnicity variable, the coding is:

```
1 Black
2 Hispanic
3 Mixed Race (Non-Hispanic)
4 Non-Black / Non-Hispanic
```

You'll want to do some data manipulations to change away from the numeric codings to more interpretable labels. 

```{r}
summary(nlsy$Income)

# Histogram
qplot(nlsy$Income)
```

The income distributing is right-skewed like one might expect.  However, as indicated in the question description, the income variable is *topcoded* at the 2% level.  More precisely,

```{r}
n.topcoded <- with(nlsy, sum(Income == max(Income, na.rm = TRUE), na.rm = TRUE))
n.topcoded
```

`r n.topcoded` of the incomes are topcoded to the maximum value of `r max(nlsy$income, na.rm = TRUE)`, which is the average value of the top `r n.topcoded` earners. You will want to think about how  to deal with this in your analysis.

<font color="#157515">
We begin our analysis by including only the variables we have shortlisted into a data-set for consistency. The trends that seem most interesting to us for further exploration in assessing what impacts income gap between genders are: Marital Status (`MaritalStatus_2017`), Highest Grade Completed, Industry, Number of Biological Children, House-hold Net Worth at 30 and a digression towards Marijuana Use. Furthermore, we would like to see whether there is an impact of physical appearance on the income gap by using Body Mass Index (BMI). 
</font>



```{r}
nlsy.recode <- nlsy %>%
  mutate(Sex = recode_factor(Sex,
                                `1` = "Men",
                                `2` = "Female",
                                `0` = "No Information"))
nlsy.recode <- select(nlsy.recode, Marijuana_1997, PerCh.Dip.20_1997, PerCh.Baby_1997, PerCh.Coll.Degree.30_1997, Sex, BDATE_M_1997, BDATE_Y_1997, Dep_F_1997, Dep_M_1997, HH_NET_WORTH_P_1997, 
 Highest.Grade_RES_DAD_1997,Highest.Grade_RES_MOM_1997,Ever_Used_Coca_1998, Disorganized_2002,
 HH_NET_WORTH_Y_2003,
 Marijuana_2005, 
  Children_HH_2007,
  Children_HH_2009, 
  Family_Income_2011, 
  Highest_Degree_Prior_2011.12,
  Children_HH_2011, 
  Highest_Grade_Completed_2011,
  Height.Inch_2011,
  Weight.Pound_2011,
  Hard.drugs.DLI_2011,
  Industry_Code_2011,
  Children_HH_2015,
  Industry_Code_2017,
  Occupation_Code_2017,
  MaritalStatus_2017,
  Income,
  HH_NET_WORTH_20,
  HH_NET_WORTH_25,
  ASSETS_FINANCIAL_25,
  ASSETS_DEBTS_20,
  HH_NET_WORTH_30,
  ASSETS_FINANCIAL,
  ASSETS_DEBTS_30
 )
```

<font color="#157515">
As we are going to be analyzing income across all our variables, we will be removing NAN values from the rows within that variable. 
</font>

```{r}
nlsy.recode <- filter(nlsy.recode, !is.na(Income)) # 
```  

<font color="#157515">
Let's assess a summary table of income across gender. 
</font>

```{r}
income.gender <- nlsy.recode %>%
  group_by(Sex) %>%
  summarize(average.income = mean(Income), 
            median.income = median(Income))
respondents <- table(nlsy.recode$Sex)
kable(income.gender, format = "markdown", digits = 3)
```
<font color="#157515">
When assessing the data via a preliminary scan, we see that the average income for men is `r income.gender$average.income[1]` and the average income for women is `r income.gender$average.income[2]`, which amounts to a difference of `r income.gender$average.income[1] - income.gender$average.income[2]`.

</font>

```{r}
ggplot(nlsy.recode, aes(Income, fill = Sex)) + geom_histogram(alpha = 0.5, position = 'identity') +
  labs (x = "Income", y = "Number of Respondents")
```

<font color="#157515">
After removing the NAN values in income. the number of female respondents is (`r respondents[2]`) compared to men (`r respondents[1]`). With the histogram, we see as the income progresses on the x-axis; men take the lead in reporting; whereas, in the lower range, women seem to be the higher number of respondents. Furthermore, we can assess that the data-set is left-skewed. It is important to note that it seems like the data is left-skewed. Let's move on to assess income gaps and gender with other variables. First up, is Marital Status.

<b>Income, Sex, and Marital Status<b>:
Would you expect a single man to earn more than a single woman? Or could a divorced man earn more than a divorced woman? Our expectation would be that marital status should not affect income depending on gender. Let's find out.
</font>

```{r}
nlsy.recode <- filter(nlsy.recode, !is.na(MaritalStatus_2017)) #removing NAN values or no responses from our analysis variable. 

nlsy.recode <-mutate(nlsy.recode, MaritalStatus_2017 = recode_factor(MaritalStatus_2017,
                                `0` = "Never-married",
                                `1` = "Married",
                                `2` = "Separated",
                                `3` = "Divorced",
                                `4` = "Widowed",
                                `NA` = "Other"))

income.gender.2<- nlsy.recode %>%
  group_by(Sex, MaritalStatus_2017) %>%
  summarize(average.income = mean(Income))

income.gender.2
```

<font color="#157515">
The simple summary showcases that the variable is worth exploring. In most instances across marital groups, the average income of men is higher than women. By running a t-test we can get more concrete information. 
</font>

```{r}

income.gap.marital <- nlsy.recode %>%
  group_by(MaritalStatus_2017) %>%
  summarize(Income.gap = mean(Income[Sex == "Men"]) -  mean(Income[Sex == "Female"]),
            lower = (t.test(x = Income[Sex == "Men"], y = Income[Sex == "Female"]))$conf.int[1],
            upper = (t.test(x = Income[Sex == "Men"], y = Income[Sex == "Female"]))$conf.int[2],
            is.significant = as.numeric(t.test(Income ~ Sex)$p.value < 0.05), .groups = "keep")

income.gap.marital
```

<font color="#157515">
The income.gap variable shows us the difference between average salaries of men and women in the specified marital groups. 
</font>

```{r}
marital.gap <- ggplot(data = income.gap.marital,aes(x = MaritalStatus_2017, y = Income.gap, fill = is.significant))
marital.gap + geom_bar(position="dodge", stat="identity") + guides(fill = guide_legend(title = "Statistically Significant")) + geom_errorbar(aes(ymin = lower, ymax = upper), width=.2, position = position_dodge(0.9)) +  theme(axis.text.x=element_text(angle = 0, vjust = 0.5)) + labs(x = "Marital Status", y = "Income Gap Between Genders")
```
<font color="#157515">
The error bars are below the 0-axis, therefore according to the data the results are not statistically significant. Running a regression model should corroborate our initial findings. First, we will run a simple regression on just income and sex; following which we will include marital status in the model
</font>

```{r}
lm.income.sex <- lm(Income ~ Sex, data = nlsy.recode)
summary(lm.income.sex)
```
<font color="#157515">
From the simple regression, it seems that on average women earn `r round(coef(lm.income.sex)["SexFemale"], 0)*-1` less than men. This seems to be statistically significant with a p-value of `r lm.income.sex$p.value`
</font>

```{r}
lm.income.sex.marital <- lm(Income ~ Sex + MaritalStatus_2017, data = nlsy.recode)
summary(lm.income.sex.marital)
```

```{r}
anova(lm.income.sex , lm.income.sex.marital)
```

<font color="#157515">
From the anova, we see that including marital status to gender has an impact on the income gap and it is significant.   
</font>

```{r}
lm.marital.interact <- update(lm.income.sex.marital, .~. + MaritalStatus_2017 * Sex)
summary(lm.marital.interact)
```
<font color="#157515">
This interactive model corresponds to differences in earnings across gender and marital status that are not accounted for by a model that assumes independent effects for Marital status and gender. For instance, here we only see that the income difference among married men and married women is $16396 less than the difference between single men and single women. However, we cannot conclude that marital status as a whole is a factor that is highly associated with the income gap between men and women.
</font>

```{r}
nlsy.recode <- data_frame(nlsy.recode)
industry.decoder <- c(rep(' AGRICULTURE, FORESTRY AND FISHERIES',121),rep('NA',79),
                    rep(' MINING',121),rep('NA',79),rep(' UTILITIES',121),
                    rep('NA',79),rep('CONSTRUCTION',1),rep('NA',299),
                    rep(' MANUFACTURING',2921),rep('NA',79),rep(' WHOLESALE TRADE',521),
                    rep('NA',79),rep(' RETAIL TRADE',1121),rep('NA',99),
                    rep(' ARTS, ENTERTAINMENT AND RECREATION SERVICES',1),
                    rep('NA',179),rep(' TRANSPORTATION AND WAREHOUSING',321),
                    rep('NA',79),rep(' INFORMATION AND COMMUNICATION',311),
                    rep('NA',89),rep(' FINANCE, INSURANCE, AND REAL ESTATE',321),
                    rep('NA',79),rep(' PROFESSIONAL AND RELATED SERVICES',521),
                    rep('NA',69),rep(' EDUCATIONAL, HEALTH, AND SOCIAL SERVICES',611),
                    rep('NA',89),rep(' ENTERTAINMENT, ACCOMODATIONS, AND FOOD SERVICES',131),
                    rep('NA',79),rep(' OTHER SERVICES',521),rep('NA',79),
                    rep(' PUBLIC ADMINISTRATION',221),rep('NA',79),
                    rep(' ACTIVE DUTY MILITARY',221),rep('NA',59),rep(' ACS SPECIAL CODES',41))
names(industry.decoder)<-170:9990

decode <- function(number) {
  if (is.na(industry.decoder[as.character(number)])) 
  {  text <- "OTHER"   }
  else {
    text <- industry.decoder[as.character(number)]   }
  return (text)}

nlsy.recode[["Industry_Code_2017"]] <- sapply(nlsy.recode[["Industry_Code_2017"]], decode)

income.industries <- nlsy.recode %>%
  group_by(Sex, Industry_Code_2017) %>%
  summarize(average.income = mean(Income, na.rm = TRUE), .groups = 'keep')

income.industries



```



```{r, fig.height = 4, fig.width = 8}
ggplot(data = income.gender.2, aes(x = MaritalStatus_2017, y = average.income, fill = Sex)) + 
  geom_bar(stat = "identity") + facet_grid(.~Sex) + coord_flip() + 
  theme(axis.text.x=element_text(angle =- 90, vjust = 0.5), legend.position = "none") + 
  labs( x = "Marital Status", y = "Average Income" )
```
```{r}
ggplot(data = nlsy.recode, aes(x = Highest_Grade_Completed_2011, y = Income , colour = Sex )) + geom_point() + stat_smooth(method = 'loess', formula = y ~ x, size = 1) + theme(axis.text.x=element_text(angle = 0, vjust = 0.5)) + xlim(0,22) + 
  labs( x = "Highest Degree Completed", y = "Average Income ($)")
```
<font color="#157515">
In the graph we can see an evolution of the income gap between men and women, depending on the highest degree completed (not using the grades as a discrete factor variable here). We don't consider in the graph the data related to non-completed undergrad studies. 
It evidences that for those who are close to 12 years of completed studies, equivalent to high school completion, the income is lower in average to those who didn't completed high school studies, and to those who have higher degree levels. It also shows that the income gap is in average lower between those who are close to 12 years of studies.
</font>


```{r}
nlsy.recode <- nlsy.recode %>%
  mutate(Marijuana_1997 = recode_factor(Marijuana_1997,
                                             `0` = "No",
                                             `1` = "Yes"))

ggplot(data = nlsy.recode, aes(x = Highest_Grade_Completed_2011, y = Income , colour = Marijuana_1997,shape = Sex)) + geom_point() + stat_smooth(method = 'loess', formula = y ~ x, size = 1) + theme(axis.text.x=element_text(angle = 0, vjust = 0.5)) + xlim(0,22) + 
  labs( x = "Highest Degree Completed", y = "Average Income ($)")
```


<font color="#157515">
This graph shows how the income in 2011 of those who had consumed marijuana by 1997, differenciate by gender across different education levels.As also evidenced in the previous graph which didn't include the marijuana consumption, the gap between those at high school level is larger to those between a higher education level.
Between 12 and 17 years of study: high school or higher, having consumer marijuana by 1997 doesn't present a difference in income for 2011, but the income gap by gender still remains.
</font>

#Interesting, let's explore how the income gap is affected with education and marital status. 
```{r}
nlsy.recode <- mutate(nlsy.recode,Highest_Grade_Completed_2011 = recode_factor(Highest_Grade_Completed_2011,
                                             `0` = "NONE",
                                             `1` = "1ST GRADE",
                                             `2` = "2ND GRADE",
                                             `3` = "3RD GRADE",
                                             `4` = "4TH GRADE",
                                             `5` = "5TH GRADE",
                                             `6` = "6TH GRADE",
                                             `7` = "7TH GRADE",
                                             `8` = "8TH GRADE",
                                             `9` = "9TH GRADE",
                                             `10` = "10TH GRADE",
                                             `11` = "11TH GRADE",
                                             `12` = "12TH GRADE",
                                             `95` = "UNGRADED",
                                             `13` = "1ST YEAR COLLEGE",
                                             `14` = "2ND YEAR COLLEGE",
                                             `15` = "3RD YEAR COLLEGE",
                                             `16` = "4TH YEAR COLLEGE",
                                             `17` = "5TH YEAR COLLEGE",
                                             `18` = "6TH YEAR COLLEGE",
                                             `19` = "7TH YEAR COLLEGE",
                                             `20` = "8TH YEAR COLLEGE OR MORE"))
income.lm.sex.marital <- lm(Income ~ Sex + MaritalStatus_2017 + Highest_Grade_Completed_2011, data = nlsy.recode)
summary(income.lm.sex.marital)
```

<font color="#157515">
As we have seen throughout the analysis, female present a lower income when compared to men. Considering `MaritalStatus_2017` and the `Highest_Grade_Completed_2011`, women in average earn $`r round(abs(coef(income.lm.sex.marital)["SexFemale"]),0)` less than men.
</font>



```{r}
income.lm.sex.degree <- lm(Income ~ Sex + Highest_Grade_Completed_2011, data = nlsy.recode)
summary(income.lm.sex.degree)
```
```{r}
anova(update(income.lm.sex.degree, . ~ . - Sex), income.lm.sex.degree)
```


```{r ,fig.height = 4, fig.width = 8}
ggplot(data = income.industries, aes(x = Industry_Code_2017, y = average.income, colour = Sex )) + 
  geom_point() + stat_smooth(method = 'lm', size = 1, fullrange = TRUE) + 
  theme(axis.text.x=element_text(angle =- 90, vjust = 0.5)) + ylab("Average Income")
```
```{r, fig.height = 4, fig.width = 8}
ggplot(data = nlsy.recode, aes(x = Income, y = Highest_Grade_Completed_2011 , colour = Industry_Code_2017, shape = Sex )) + geom_point() + stat_smooth(method = 'loess', formula = y ~ x, size = 1) + theme(axis.text.x=element_text(angle =- 90, vjust = 0.5)) + ylim(0,22)
```
```{r}
income.lm.grade.ind.sex <- lm(Income ~ Highest_Grade_Completed_2011 + Industry_Code_2017 + Sex, data = nlsy.recode)
summary(income.lm.grade.ind.sex)
```
```{r}
confint(income.lm.grade.ind.sex, parm = "Industry_Code_2017")
confint(income.lm.grade.ind.sex, parm = "Highest_Grade_Completed_2011")
confint(income.lm.grade.ind.sex, parm = "Sex")
```
```{r}
income.lm.grade.sex <- lm(Income ~ Highest_Grade_Completed_2011 + Sex, data = nlsy.recode)
summary(income.lm.grade.sex)
```

```{r}
income.lm.sex <- lm(Income ~ Sex, data = nlsy.recode)
summary(income.lm.sex)
```
```{r}
income.lm.bio.child.sex <- lm(Income ~ Children_HH_2009 + Sex, data = nlsy.recode)
summary(income.lm.bio.child.sex)
```
```{r}
anova(update(income.lm.grade.ind.sex, . ~ . - Sex), income.lm.grade.ind.sex)
```

<font color="#157515">
Income, sex and number of children. Would you expect income to increase regardless of gender as the family size increases? On the flip side, we want to explore whether a larger family size is correlated with a lack of education leading to a lower income across the board. If women are educated, they might have fewer children and that might contribute to their higher income. In order to do so, we have chosen the "biological children in household data" from 2009. We checked the data and it seemed to be replicated for all the years. We will run simple summaries and check for associations, next we will run a simple regression model to check for significance, and finally explore if there are any interactions between highest grade completed, number of children, sex and income.  

</font>

```{r}
num.bio.child <- nlsy.recode %>%
  group_by(Children_HH_2009, Sex) %>%
  summarize(average.income = mean(Income, na.rm = TRUE))

num.bio.child <- filter(num.bio.child, !is.na(average.income),!is.na(Children_HH_2009))
num.bio.child
```
<font color="#157515">
Running a simple summary shows us that, across the board, as the number of children increase womens' incomes are less than men. With the outlier response of $2,500 from Men with six children. This is something we can explore further.    
</font>
```{r, fig.height = 4, fig.width = 9}
ggplot(data = num.bio.child, aes(x = Children_HH_2009, y = average.income, fill = Sex)) + 
  geom_bar(position="dodge", stat="identity") +
  theme(axis.text.x=element_text(angle = 0, vjust = 0.5)) + 
  labs( x = "Number of Biological Children", y = "Average Income ($)")
```
```{r}
#running a simple regression model for Income, Biological Children and Sex 
income.lm.bio.child.sex <- lm(Income ~ Children_HH_2009 + Sex, data = nlsy.recode)
summary(income.lm.bio.child.sex)
```
<font color="#157515">
From the simple regression, it seems that on average women earn `r round(coef(income.lm.bio.child.sex)["SexFemale"], 0)*-1` less than men given that the number of children are the same. This seems to be statistically significant with a p-value of `r income.lm.bio.child.sex$p.value`. 
</font>

<font color="#157515">
Interesting, could adding the highest degree earned to the existing model exacerbate or mitigate the income gap between men and women?
</font>

```{r}
income.lm.bio.child.sex.edu <- lm(Income ~ Children_HH_2009 + Sex + Highest_Grade_Completed_2011, data = nlsy.recode)
summary(income.lm.bio.child.sex.edu)
```

```{r}
used.coca <- nlsy.recode %>%
  group_by(Ever_Used_Coca_1998, Sex) %>%
  summarize(mean = mean(Income),
            lower = t.test(Income, na.rm = TRUE)$conf.int[1],
            upper = t.test(Income, na.rm = TRUE)$conf.int[2],
            is.signif = as_factor(
              if (lower<0 & upper>0){
                is.signif = 0
              } else {
                is.signif = 1
              }
            ), .groups = 'keep')
used.coca

```
```{r}
num.bio.child <- nlsy.recode %>%  
  group_by(Children_HH_2009, Sex) %>%
  summarize(mean = mean(Income[Sex == "Men"]) -  mean(Income[Sex == "Female"]),
            lower = t.test(x = Income[Sex == "Men"],
                            y = Income[Sex == "Female"])$conf.int[1],
            upper = t.test(x = Income[Sex == "Men"],
                            y = Income[Sex == "Female"])$conf.int[2], .groups = "keep")
```

```{r}
education.indicator <- filter(nlsy.recode, !(nlsy.education$Highest_Grade_Completed_2011 == "NONE"),
                              !(nlsy.education$Highest_Grade_Completed_2011 == "1ST GRADE"),
                              !(nlsy.education$Highest_Grade_Completed_2011 == "2ND GRADE"),
                              !(nlsy.education$Highest_Grade_Completed_2011 == "3RD GRADE"),
                              !(nlsy.education$Highest_Grade_Completed_2011 == "4TH GRADE"),
                              !(nlsy.education$Highest_Grade_Completed_2011 == "5TH GRADE"),
                              !(nlsy.education$Highest_Grade_Completed_2011 == "6TH GRADE"),
                              !(nlsy.education$Highest_Grade_Completed_2011 == "7TH GRADE"),
                              !(nlsy.education$Highest_Grade_Completed_2011 == "UNGRADED"),
                              !is.na(nlsy.education$Highest_Grade_Completed_2011),
                              !is.na(nlsy.education$Income))
ed.indicator <- education.indicator %>%
  group_by(Highest_Grade_Completed_2011) %>%
  summarize(mean = mean(Income),
            lower = t.test(Income)$conf.int[1],
            upper = t.test(Income)$conf.int[2],
            is.signif = as_factor(
              if (lower<0 & upper>0){
                is.signif = 0
              } else {
                is.signif = 1
              }
              ), .groups = 'keep')

```


```{r}
education.indicator <- mutate(ed.indicator,
                     Highest_Grade_Completed_2011 = factor(Highest_Grade_Completed_2011,
                     levels = c("8TH GRADE", "9TH GRADE", "10TH GRADE",
                     "11TH GRADE", "12TH GRADE", "1ST YEAR COLLEGE", "2ND YEAR COLLEGE",
                     "3RD YEAR COLLEGE", "4TH YEAR COLLEGE", "5TH YEAR COLLEGE", "6TH YEAR COLLEGE",
                     "7TH YEAR COLLEGE", "8TH YEAR COLLEGE OR MORE")))
nlsy.education.plot <- ggplot(data = ed.indicator,
                              aes(x = Highest_Grade_Completed_2011, y = mean,
                                  fill = is.signif))
nlsy.education.plot + geom_bar(position="dodge", stat="identity") +
  guides(fill = guide_legend(title = "Statistically Significant")) +
  geom_errorbar(aes(ymin = lower, ymax = upper), 
                width=.2, position = position_dodge(0.9)) +
  theme(axis.text.x=element_text(angle = 90, vjust = 0.5))
```
